---
title: "Effect of management analysis"
author: "John R. Foster"
format: html
execute: 
  warning: false
  error: false
  echo: false
---

```{r}
#| label: setup
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
```

This document will run the analysis for the manuscript

First, make sure everything is up to date:

```{r}
#| label: run targets pipeline
#| echo: true
tar_make()
```

### Population trajectory without management

```{r}
#| label: fig-trajectory
#| fig-cap: "Figure shows the change in median wild pig density relative to the initial median density at week 0 over a one-year period. The boxplots represent the median and interquartile range (IQR) of these changes, where the whiskers extend plus or minus 1.5 times the IQR with outliers removed. The Without management scenario (orange) is the result of the boostrapping procedure (see methods section ##), and the With managment scenario (purple) compares median posterior density estimates. The rows represent the density category at week 0, where Low is less than 2 pigs/km2, Medium is between 2-6 pigs/km2, and High is greater than 6 pigs/km2. Note the different y-axis scale for each initial density category."

tar_read("plot_managment")

```

```{r}
#| label: bootstap analysis
df <- tar_read("bootstrap_mgmt")

md <- max(df$delta_time)

cat_medians <- df |>
  filter(
    delta_time == md
  ) |>
  group_by(st_name, density_cat, mgmt) |>
  summarise(med = median(delta_density), 2) |>
  ungroup()

get_median_changes <- function(df, dc, mc) {
  df |>
    filter(
      density_cat == dc,
      mgmt == mc
    ) |>
    filter(med == min(med) | med == max(med)) |>
    arrange(med) |>
    mutate(med = round(med, 2))
}

tmp <- cat_medians |> get_median_changes("High", "Without")
high_wo_states <- tmp$st_name
high_wo_meds <- tmp$med

tmp <- cat_medians |> get_median_changes("High", "With")
high_with_states <- tmp$st_name
high_with_meds <- tmp$med

tmp <- cat_medians |> get_median_changes("Low", "Without")
low_wo_states <- tmp$st_name
low_wo_meds <- tmp$med

tmp <- cat_medians |> get_median_changes("Low", "With")
low_with_states <- tmp$st_name
low_with_meds <- tmp$med

```

The effect of pig removals was more evident when the initial density was high (@fig-trajectory, bottom row). For example, when initial density was high (i.e. \> 6 pigs/km2) and management did not occur, the range in the change in density over one year was predicted to have a median increase between `r high_wo_meds[1]` and `r high_wo_meds[2]` pigs/km2 for `r high_wo_states[1]` and `r high_wo_states[2]`, respectively. By contrast, when management took place in high density areas, the range in the change in density over one year was observed with a median decrease between `r high_with_meds[1]` and `r high_with_meds[2]` pigs/km2 for `r high_with_states[1]` and `r high_with_states[2]`, respectively.

When the initial density was low (i.e. \< 2 pigs/km2), the effect of management over a one year period was mixed (@fig-trajectory, top row). For example, when management did not occur, the range in the change in density over one year was predicted to have a median change between `r low_wo_meds[1]` and `r low_wo_meds[2]` pigs/km2 for `r low_wo_states[1]` and `r low_wo_states[2]`, respectively. By contrast, when management took place in low density areas, the range in the change in density over one year was observed with a median between `r low_with_meds[1]` and `r low_with_meds[2]` pigs/km2 for `r low_with_states[1]` and `r low_with_states[2]`, respectively.

### Population growth rate

```{r}
#| label: fig-lambda
#| fig-cap: "Figure shows the average population growth rate at each property within each state. The number of properties (n) within each state are on the y-axis. A growth rate greater than one indicates that a property's population is growing. Values of zero represent properties that went extinct between the first and second observed primary periods. Boxplots represent the median and interquartile range (IQR), where the whiskers extend plus or minus 1.5 times the IQR."

tar_read("plot_lambda_hat_by_state")

```

```{r}
#| label: lambda analysis

df <- tar_read("mu_lambda")

all_prop_med <- round(median(df$mu), 2)
all_prop_low <- round(quantile(df$mu, 0.05), 2)
all_prop_high <- round(quantile(df$mu, 0.95), 2)

n_props <- nrow(df)
n_declining <- df |>
  filter(mu < 1) |>
  nrow()
percent_declining <- round(n_declining / n_props * 100)

lambda_summary <- df |>
  group_by(st_name) |>
  summarise(
    n = n(),
    med = median(mu),
    sd = sd(mu),
    q1 = quantile(mu, 0.25),
    q3 = quantile(mu, 0.75)
  ) |>
  arrange(med)

low <- lambda_summary |>
  filter(med == min(med))
state_low <- low$st_name
low_lambda <- round(low$med, 2)

high <- lambda_summary |>
  filter(med == max(med))
state_high <- high$st_name
high_lambda <- round(high$med, 2)


low_sd <- lambda_summary |>
  filter(sd == min(sd, na.rm = TRUE))
state_low_sd <- low_sd$st_name
low_lambda_sd <- round(low_sd$sd, 2)

high_sd <- lambda_summary |>
  filter(sd == max(sd, na.rm = TRUE))
state_high_sd <- high_sd$st_name
high_lambda_sd <- round(high_sd$sd, 2)

n_under_1 <- lambda_summary |>
  filter(med < 1) |>
  nrow()

```

Individual properties exhibited a range of average population growth rates. Across all properties the median population growth rate per 28-day period was `r all_prop_med` (90% CI: `r all_prop_low` - `r all_prop_high`). There was also variation within and across states (@fig-lambda). For example, the median growth rate per 28-day period ranged from `r low_lambda` to `r high_lambda` for `r state_low` and `r state_high`, respectively. The standard deviation of growth rates within each state ranged from `r low_lambda_sd` to `r high_lambda_sd` for `r state_low_sd` and `r state_high_sd`, respectively. There were `r n_under_1` states where at least half of their properties had a declining population (growth rate under one), and `r percent_declining`% of all properties (n=`r n_declining`) had a declining population.

### Proportion removed

```{r}
#| label: fig-removals
#| fig-cap: "The boxplots in panel A are drawn around the mean proportion of the population removed by method at each property. Boxplots represent the median and interquartile range (IQR), where the whiskers extend plus or minus 1.5 times the IQR with outliers shown. Panel B shows the theoretical proportion of the population that needs to be removed given the management goal and population growth rate (equation ##)."

tar_read("plot_prop_take")

```

```{r}
#| label: proportion take
df_actual <- tar_read("actual_percent_take")
df_theoretical <- tar_read("theoretical_percent_take")

by_state <- function(df) {
  df |>
    group_by(method, st_name) |>
    summarise(
      med = median(mu),
      low = quantile(mu, 0.05),
      high = quantile(mu, 0.95)
    ) |>
    ungroup()
}

a_by_state <- by_state(df_actual)

grid_search <- expand_grid(
  lambda = seq(0.5, 1.5, by = 0.001),
  goal = c(0, 0.5) # 0 is no change, 1 would be elimination
) |>
  filter(goal + lambda >= 1) |>
  mutate(needed = -1 * (((1 - goal) - lambda) / lambda))


tx_heli <- a_by_state |>
  filter(method == "Helicopter", st_name == "TEXAS")

tx_heli_med <- round(tx_heli$med, 2)
tx_heli_low <- round(tx_heli$low, 2)
tx_heli_high <- round(tx_heli$high, 2)

get_example <- function(g) {
  grid_search |>
    mutate(needed = round(needed, 2)) |>
    filter(
      goal == g,
      needed == tx_heli_med
    ) |>
    pull(lambda) |>
    mean() |>
    round(2)
}

tx_0 <- get_example(0)
tx_0.5 <- get_example(0.5)


pnorm(0.67, 0.978, 0.323)
pnorm(0.784, 0.978, 0.323)
```

Here, we focus on the proportion of the population removed for properties that have growing populations because we are interested in how removals might be used to reach a management goal (i.e. a 50% reduction in abundance), and what amount of the population that needs to be removed to reach that goal. This only makes sense for growing populations

To interpret @fig-removals, we start with a specific example to explain how to interpret the data, and then later we report the findings at large. For the specific example, let's look at Texas. The median proportion of the population removed in a single primary period, by helicopters (from @fig-removals A) was `r tx_heli_med` (90% CI `r tx_heli_low` - `r tx_heli_high`). If the population was growing, and the management goal was to at maintain the population at its current level, then the population growth rate would have to be under `r tx_0` to achieve this goal (@fig-removals B). If, however, the goal was to reduce the population by half, then the growth rate would have to be under `r tx_0.5`.